Understanding Regional and Temporal Variations in Food Prices: Insights from US Consumer Price Index and Economic Trends

### 2024-11-26 paper/paper.qmd
outcome variable summary

@@

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Talk more about it.

Talk way more about it.

>>>>

library(knitr)
library(ggplot2)

@@

## Outcome variable

The response variable, unitValue, represents the weighted mean price per 100 grams of a food item within a specific food category, geographic region, and time period. In this analysis, the variable is log-transformed to create unitValue_lg, which helps stabilize variance, improve normality, and facilitate the interpretation of model coefficients in percentage terms. The log transformation allows us to understand the proportional changes in price rather than absolute changes, which is particularly useful for identifying relative price dynamics across categories and regions.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: out-summ-stats
#| tbl-cap: "Summary Statistics for log of Unit Value"

# Summary statistics for outcome variable
summary_stats_outcome <- analysis_data %>%
  summarise(
    mean = mean(unitValue_lg, na.rm = TRUE),
    median = median(unitValue_lg, na.rm = TRUE),
    sd = sd(unitValue_lg, na.rm = TRUE),
    min = min(unitValue_lg, na.rm = TRUE),
    max = max(unitValue_lg, na.rm = TRUE)
)

kable(summary_stats_outcome, )

The transformed outcome variable, unitValue_lg, represents the log-transformed weighted mean price per 100 grams of food items, with a mean of -0.775 and a median of -0.777, indicating a near-symmetric distribution. The standard deviation of 0.746 reflects moderate variability in prices, while the range, spanning from a minimum of -3.411 to a maximum of 1.191, highlights significant disparities in food costs across categories, regions, and time periods. The log transformation normalizes the data, reducing skewness and ensuring that extreme values in the original price variable do not disproportionately influence the analysis.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: out-summ-stats
#| tbl-cap: "Summary Statistics for log of Unit Value"

# Histogram for unitValue_lg

ggplot(analysis_data, aes(x = unitValue_lg)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of Log Unit Value", x = "Log Unit Value", y = "Frequency") +
  theme_minimal()

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: out-summ-stats
#| tbl-cap: "Summary Statistics for log of Unit Value"

filtered_data <- analysis_data %>%
  filter(grepl("Census Region", region, ignore.case = TRUE))

# Boxplot of unitValue_lg by region
ggplot(filtered_data, aes(x = region, y = unitValue_lg)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.7) +
  labs(title = "Log Unit Value by Region", x = "Region", y = "Log Unit Value") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
